#Better Data Analysis with R
##Example 2 - Repair Statistics
###*Eric Koziol*

```{r}
library(lattice)
library(ggplot2)
haunches <- read.csv("Examples/Example_2_Data/Haunches.csv", header=FALSE)
names(haunches) <- c("id", "haunchid", "bent", "span", "haunch", "currentRepair", "specifiedRepair", "differentStates", "itemsComplete", "Stage", "inRepairSet", "inspected")

haunchesInRepairs <- haunches[haunches$inRepairSet == 1,]
```

#Problem
WJE is overseeing repairs of elevated rails and needs to understand the status of areas that are being completed.  Additionally, better visualization of repair locations is needed.


#Solution
WJE created a digital inspection system for field inspectors to track the completion of repairs.  While the data collection system is robust, the reporting mechanisms are not.  The data is routinely exported into R for digestion of analysis and tracking of progress.

```{r, fig.height=100}
#names(septaRepairs)
histogram(~currentRepair|factor(bent), layout=c(2,52), data = haunches[with(haunches, order(-bent)),], labels=TRUE, type="count")


```

```{r, fig.height=300, message = FALSE, warnings = FALSE}
ggplot(data=haunchesInRepairs, aes(x=currentRepair, fill=currentRepair)) + geom_histogram() + geom_text(aes(y=-0.5, label=currentRepair)) + stat_bin(binwidth=1, geom="text", aes(label=..count..), vjust=-0.5) + facet_wrap(~bent, ncol=2)
```

```{r}
library(reshape2)
library(scales)
haunchesMelt <- melt(haunchesInRepairs[,c("bent", "Stage")], id.vars="bent")
head(haunchesMelt,10)
str(haunchesMelt)
haunchesMelt$value <- as.factor(haunchesMelt$value)
str(haunchesMelt)
```