
#Better Data Analysis with R
##Plotting
###*Eric Koziol*


#Setup
Before we get started please ensure that the following packages are installed and running.

```{r, eval=FALSE}
install.packages("ggplot2")
install.packages("lattice")
install.packages("AppliedPredictiveModeling")
```
```{r}
library(ggplot2)
library(lattice)
library(AppliedPredictiveModeling)
```

The Applied Predictive Modeling package contains a concrete dataset that we can experiment with.  I thought it would be good to use a dataset that our company actually deals with.  Let's load it now.

```{r}
data(concrete)
head(mixtures,10)
```

In order to view the top ten rows to get a feel for the data we can use the **head(*dataframe*,*number of rows*)** command.


#Plotting systems in R

There are three main plotting systems in R - Base, Lattice and ggplot2.  The Base plotting system runs out of the box on R, while the lattice and ggplot2 plotting systems need to be loaded with the appropriate library - **libray(lattice)** and **library(ggplot2)**.

